---
title: "NBA Players Statistics Analysis"
subtitle: "Data Preparation"
author: 
  - Ngoc Nguyen

date: "12/07/2022"
output: 
  html_document:
    theme: cosmo
    toc: true
    number_sections: true
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                     # results = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      #fig.show = 'hide',
                      fig.align='left')
```

# Data Preparation

The data used for the analysis constitutes of NBA player data for the season from the 2018-2019 season to the 2021-2022 season.This combined data for the season is sourced from <https://www.nbastuffer.com>.

This section presents the preparation of the data for further analysis and modeling. The following actions have been performed on the data:

- Load the Dataset

- Rename the columns 

- Drop columns that are not usefull for the analysis

- Save the data as .RData

- Extract codebook 

- Save the codebook as html or pdf


# Load the required Libraries

```{r Load Libraries}

source('config/00_dependencies.R')

```

# Load Data

```{r Load Data}

nba_data <- read_csv("data/nba_data.csv")

nba_labels <- read_csv("data/nba_data_labels.csv")

```

# Rename Variables
```{r Variable Names}

colnames(nba_data)
colnames(nba_labels)

```

# Rename Variables
```{r Rename Variables}

colnames(nba_data) <- nba_labels$name

colnames(nba_data)

```

# Drop Columns 

The following columns are dropped from the dataset:

-`rank`. the `rank` row has no data.

- `full_name`. the player name is dropped and just remain with the observations.

- `season`. this is a categorical value representing the seasons for which the data was collected. We will not include this as part of the predictor variables. 

```{r Drop Columns}

cols <- c('rank', 'full_name', 'season')

nba_data_select <- select(nba_data, -cols)

colnames(nba_data_select)

```
# Save to .RData File
```{r Save .Rdata File}

nba_dataset <- data.frame(nba_data_select)

save(nba_dataset, file="data/nba_dataset.RData")

```


# Create a Codebook
```{r Create a Codebook}

code <- colnames(nba_data_select)

code_book <- data.frame(code)

code_book$label <- nba_labels$label[3:29]

code_book <- code_book %>%
  mutate(type = map_chr(nba_data_select, typeof),
         mean = round(map_dbl(nba_data_select, mean, na.rm=T),2), 
         prop_miss = round(map_dbl(nba_data_select, function(x) mean(is.na(x))),4)) 
  
```


# Save the Code Book
```{r save Code Book}

kable(code_book, "html", caption="Code Book") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  cat(., file = "outputs/code_book.html")

```



